---
title: "USA extreme weather events exploration"
author: "Marko"
date: "11/19/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.width = 12, fig.height = 8)
```

# Synopsis



# Data Processing

## Import raw data

Firs we need to import some libraries to execute the analysis:
```{r firststep, message=FALSE}
rm(list = ls())
graphics.off()

# Load R packages
packages <- c("dplyr",   # list of packages to load
              "ggplot2",  
              "lubridate", 
              "stringr", 
              "cowplot", 
              "data.table",
              "janitor",
              "kableExtra") 

package.check <- lapply( # load or install & load list of packages
  packages,
  FUN = function(x) {
    if (!require(x, character.only = TRUE)) {
      install.packages(x, dependencies = TRUE)
      library(x, character.only = TRUE)
    }
  }
) 
rm(packages, package.check)

```


Raw data is downloaded here:
```{r import, cache=TRUE}
# Download raw csv file if necessary
if(!file.exists("repdata_data_StormData.csv.bz2")){
  download.file(url = "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", 
                destfile = "repdata_data_StormData.csv.bz2") 
  }

# Import file
df.raw <- fread(input = "repdata_data_StormData.csv.bz2")
```


## Initial data inspection

Dimension (number of observations & number of variables) of the dataset:
```{r dim}
dim(df.raw)
```

Structure of given data frame:
```{r str}
str(df.raw)
```

Sneak peak into first few rows of the data:
```{r head}
head(df.raw) %>% 
  kbl() %>% 
  kable_paper() %>%
  scroll_box(width = "100%", height = "250px")
```




## Data wrangling

We need to clean the initial data set a bit in order to answer our questions.

First we will clean columns names and set proper variable types;

```{r wrangling}
# clean column names
df.clean <- df.raw %>% 
  clean_names  

# column type conversion & create new columns
df.clean <- df.clean %>% 
  mutate(bgn_date = mdy_hms(bgn_date), # convert to date
         bgn_datetime = paste0(as.character(bgn_date), " ",  # build date + time column
                               str_sub(bgn_time, start = 1, end = 2), ":",
                               str_sub(bgn_time, start = 3, end = 4)),
         bgn_datetime = ymd_hm(bgn_datetime) # convert to date time object
         )
  
  
  
```


# Results